---
title: "Technical Reference: Setup, Domain, and Business Rules"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Technical Reference: Setup, Domain, and Business Rules}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Part 1: Environment and Prerequisites

## R Package Dependencies

afbuildr imports from 9 packages. Most install from CRAN:
```{r}
install.packages(c(
  "data.tree", "dplyr", "lubridate", "purrr",
  "sf", "stringr", "tidyr"
))
```

The `piwebapi` package provides the PI Web API client and is **not on
CRAN**. Install it from its internal source:
```{r}
# Replace with your actual installation method
# e.g., remotes::install_github("org/piwebapi")
```

`stats` and `utils` ship with base R.

## External Functions Not Defined in This Package

Two functions are called by the pipeline but are **not defined** in
afbuildr. They must be available in the R session before running:
- `load_gis_ps()` --- loads pump station features from the GIS system.
  Returns an `sf` data frame with at least columns `FAC_STATE`,
  `CMMS_ASSET_ID`, `SUBTYPE`, `STREET_NAME`, `STREET_NUMBER`, `PI_TAG`.
- `load_paths()` --- sets working directory paths. Called at the start of
  `validate_gis()`.
- `get_webid()` --- provided by the `piwebapi` package. Looks up the web
  ID for a PI tag path. afbuildr wraps this as `safe_get_webid()` via
  `purrr::possibly()` to return `NULL` instead of erroring on missing tags.

## PI Web API Access

The `piwebapi` package must be configured with:

- **Server URL** for the PI Web API endpoint
- **Authentication credentials** (typically Windows integrated or basic
  auth)
- **Network access** to the PI Data Archive (may require VPN or on-site
  connection)

Verify connectivity before running the pipeline:
```{r}
# Should return a web ID string, not NULL
safe_get_webid("wwl:north:mai100_P1RUN")
```

If this returns `NULL` for a tag you know exists, the PI Web API connection
is not working.

## Required Data Files

Two data files must be present in the `data/` directory before running
the pipeline:

| File | Used by | Contents |
|------|---------|----------|
| `data/ls_by_plant.RData` | `get_ls_by_plant()` (Step 2) | Lookup table mapping `CMMS_ASSET` to `BASIN` |
| `data/dia_by_ls.RData` | `get_dia_by_ls()` (Step 2) | Lookup table mapping `CMMS_ASSET_ID` to pipe `Dia` (diameter) |

These files are not generated by the package --- they come from external
sources. The `get_ls_by_plant()` and `get_dia_by_ls()` functions are
marked TODO for replacement with computed lookups from GIS/CMMS.

## Data Flow Between Steps

```
Step 1: validate_gis()
  Reads:  GIS (via load_gis_ps), PI Web API
  Writes: data/pi_tag_master_list.csv
            |
            v
Step 2: build_af()
  Reads:  data/pi_tag_master_list.csv, PI Web API, SCADA history,
          data/ls_by_plant.RData, data/dia_by_ls.RData
  Writes: data/pi_af_import.csv
            |
            v
Step 3: export_af()
  Reads:  data/pi_af_import.csv
  Writes: xlsx/pi_af_loader.csv
```

### Intermediate File Schemas

**`data/pi_tag_master_list.csv`** (output of Step 1):

| Column | Type | Example |
|--------|------|---------|
| `CMMS_ASSET_ID` | character | `"LS001"` |
| `PI_TAG` | character | `"WWL:NORTH:MAI100"` |
| `STREET_NAME` | character | `"MAIN"` |
| `STREET_NUMBER` | character | `"100"` |

**`data/pi_af_import.csv`** (output of Step 2):

| Column | Type | Example |
|--------|------|---------|
| `BASIN` | character | `"BUCKMAN"` |
| `STATION_NAME` | character | `"MAIN - 100"` |
| `CMMS_ASSET_ID` | character | `"LS001"` |
| `PI_TAG` | character | `"NORTH:MAI100"` (WWL: prefix removed) |
| `Dia` | numeric | `50.3` |
| `TEMPLATE` | character | `"_cte_leadsp"` or `"_cte_lead"` |

**`xlsx/pi_af_loader.csv`** (output of Step 3):

| Column | Type | Description |
|--------|------|-------------|
| `Selected(x)` | character | Always `"x"` |
| `Parent` | character | AF hierarchy path, backslash-separated |
| `Name` | character | Element or attribute name |
| `ObjectType` | character | `"Element"` or `"Attribute"` |
| `Template` | character | AF template name |
| `AttributeValue` | character | Attribute value (if applicable) |


# Part 2: Domain Glossary

## PI System Terminology

- **PI Data Archive** --- OSIsoft time-series database (the "historian")
  that stores SCADA data.
- **PI Web API** --- RESTful API for querying PI data. Used by this package
  to validate whether a tag exists and to retrieve historical values.
- **Web ID** --- Unique identifier for a PI object. `get_webid()` returns
  this; a `NULL` result means the tag does not exist.
- **Asset Framework (AF)** --- Hierarchical model in PI that organizes
  assets (stations, pumps) into a tree structure. The CSV loader imports
  elements and attributes into this hierarchy.
- **AF Template** --- Predefined structure applied to an AF element.
  Templates used by this package:
  - `_Liftstation` --- station-level template
  - `_cte_lead` --- constant-time-estimate flow calculation (no lead
    setpoint)
  - `_cte_leadsp` --- constant-time-estimate flow calculation (with lead
    setpoint)
  - `_hrt` --- hydraulic retention time calculation

## SCADA Tag Types

Each pump station can have up to 4 pumps, and each pump has 4 tag types:

| Suffix | Full Name | What It Measures |
|--------|-----------|------------------|
| `RUN` | Run status | Binary (0/1) indicating whether the pump is running |
| `FRQ` | Frequency | Motor electrical frequency in Hz (indicates VFD operation) |
| `RSPD` | Running speed | Motor running speed (indicates VFD operation) |
| `CSPD` | Command speed | Motor commanded speed (indicates VFD operation) |

A full tag path for Pump 1's run status at a station looks like:
`wwl:north:mai100_P1RUN`

- **Lead setpoint (`_WLLEADSP`)** --- Wet well level setpoint that triggers
  the lead pump. Its presence indicates the station uses level-based pump
  control, which determines the AF template choice.

## Pump Station Terminology

- **Lift station (LS)** --- A wastewater pump station that lifts sewage
  from a lower elevation to a higher one. Identified by `CMMS_ASSET_ID`
  starting with `"LS"`.
- **Booster pump station (BPS)** --- A station that boosts pressure in the
  force main. Identified by `CMMS_ASSET_ID` starting with `"BPS"`.
  Booster stations use suffix `"B"` in the tag naming convention.
- **VFD (Variable Frequency Drive)** --- A motor controller that adjusts
  pump speed. VFD-controlled pumps have active FRQ, RSPD, or CSPD tags
  (values that vary over time), as opposed to constant-speed pumps that
  only have RUN tags.
- **Lead/lag configuration** --- Standard pump control where one pump
  (lead) runs primarily and a second (lag) activates when demand exceeds
  the lead pump's capacity. The CTE flow estimation model assumes this
  configuration, which is why stations with more than 2 pumps are excluded.
- **Force main** --- The pressurized pipe downstream of the pump station.
  Its diameter (`Dia`) is used in flow calculations. Measured as
  cross-sectional area in square inches.
- **Wetwell** --- The underground chamber where wastewater collects before
  being pumped. Wetwell area is derived from the force main diameter and
  stored in the AF hierarchy.

## Tag Naming Convention

PI tags follow the format:

```
wwl:{grid}:{street_code}{address}{suffix}_P{pump}{type}
```

| Component | Values | Example |
|-----------|--------|---------|
| `wwl` | Literal prefix (wastewater lift) | `wwl` |
| `grid` | `north`, `east`, `south`, `west`, `nassau`, `stjohns` | `north` |
| `street_code` | First 3 characters of street name (spaces removed) | `mai` (from "MAIN") |
| `address` | Street number | `100` |
| `suffix` | `""` (standard), `"B"` (booster), `"-1"` (new station) | `B` |
| `P{pump}` | Pump number 1--4 | `P1` |
| `type` | `RUN`, `FRQ`, `RSPD`, `CSPD` | `RUN` |

The base tag (without pump/type suffix) always ends with `_`:
`wwl:north:mai100_`

## Grid Regions

The six grid regions correspond to geographic service areas:

| Grid | Area |
|------|------|
| `north` | North service area |
| `east` | East service area |
| `south` | South service area |
| `west` | West service area |
| `nassau` | Nassau County |
| `stjohns` | St. Johns County |

## GIS Field Codes

| Field | Meaning | Key Values |
|-------|---------|------------|
| `FAC_STATE` | Facility operational state | `"INS"` = in-service |
| `SUBTYPE` | Facility subtype | `1` = lift station |
| `CMMS_ASSET_ID` | CMMS (maintenance system) identifier | `"LS001"`, `"BPS002"` |


# Part 3: Business Rules and Thresholds

## Step 1: Tag Classification Rules

`classify_tags()` splits tags into categories based on format, then
validates each category against the PI Web API.

### Classification categories

1. **NA tags** --- `PI_TAG` is `NA` in GIS. The station has no tag
   recorded.
2. **NONE tags** --- `PI_TAG` contains `"NONE"`. The station has been
   explicitly marked as having no SCADA connection. These are excluded
   from further processing.
3. **Missing underscore** --- `PI_TAG` does not end with `_`. This is a
   data entry formatting error. The tag is cleaned (spaces, `-*`, and
   line breaks removed, then `_` appended) and re-validated.
   - *Valid Otherwise*: the cleaned tag resolves in PI Web API.
   - *Bad Tag*: the cleaned tag still does not resolve.
4. **Correct format** --- `PI_TAG` ends with `_` as expected.
   - *OK*: the tag resolves in PI Web API.
   - *Bad Tag*: the tag does not resolve despite correct format.

### Validation method

Every tag is validated by appending `P1RUN` and calling
`safe_get_webid()`. If Pump 1's run status tag exists, the base tag is
considered valid. This is a proxy --- it assumes every station has at
least a P1RUN tag.

## Step 1: Tag Guessing Rules

`guess_tag_name()` attempts to reconstruct a tag name from the station's
street address when the existing tag is missing or invalid.

### Candidate generation

For each asset, candidates are generated as:
```
wwl:{grid}:{first 3 chars of street}{street number}{suffix}_P1RUN
```

This produces up to **18 candidates** per asset (6 grids x 3 suffixes).
The search stops at the first match to minimize API calls.

### Suffix meanings

| Suffix | Meaning | When used |
|--------|---------|-----------|
| `""` (empty) | Standard station | Tried first |
| `"B"` | Booster pump station | Tried second |
| `"-1"` | New station placeholder | Tried third |

### Search order

Suffixes are tried in the order above. Within each suffix, all 6 grids
are tried. This means the empty suffix (standard) is tried for all grids
before moving to the `"B"` suffix, biasing toward standard stations.

## Step 2: VFD Detection Thresholds

`check_multiple_pumps_or_vfd_control()` analyzes 7 days of SCADA history
to classify stations.

### Outlier threshold: values > 60

SCADA values above 60 are set to `NA`. For RUN tags (binary 0/1), any
value above 60 is clearly a sensor error. For frequency tags, motor
frequency should be at or below 60 Hz (the standard US power frequency).
Values above 60 across all tag types indicate data quality issues.

### VFD indicator: speed/frequency > 35

CSPD, RSPD, and FRQ tag values above 35 indicate variable frequency drive
operation. A constant-speed pump would show either 0 (off) or ~60 Hz
(on). Values between 35 and 60 indicate the drive is actively modulating
speed. The threshold of 35 is set conservatively below the 60 Hz
operating point to capture pumps running at reduced speed.

### Activity detection: standard deviation > 0

A tag is considered active only if its standard deviation is greater than
0 over the 7-day window. Tags with `sd == 0` are either flatlined
(sensor fault) or the pump never ran during the sample period.

### Pump count detection

The pump number is extracted from the tag name (the digit after `P` in
`_P1RUN`, `_P2RUN`, etc.). The maximum pump number per station determines
the pump count. Stations with `pump > 2` are flagged because the CTE
flow model assumes a lead/lag (2-pump) configuration.

## Step 2: Template Assignment

Each station is assigned an AF template based on whether a lead setpoint
tag exists:

| `_WLLEADSP` tag exists? | Template assigned | Meaning |
|--------------------------|-------------------|---------|
| Yes | `_cte_leadsp` | Station uses level-based lead pump control with a configurable setpoint |
| No | `_cte_lead` | Station uses lead pump control without a setpoint tag in PI |

## Step 2: Pipe Diameter Normalization

Force main diameter is stored as cross-sectional area in square inches
(`Dia` column). Two rules are applied:

| Condition | Rule | Value | Rationale |
|-----------|------|-------|-----------|
| `Dia` is `NA` | Use default | 50.3 sq in (~8" pipe) | 8-inch is the most common force main diameter in the system |
| `Dia < MIN_WETWELL_AREA` | Use minimum | 19.6 sq in (5" pipe) | 5 inches is the smallest force main in the collection system |

## Step 2: Station Exclusions

Stations are excluded from the AF import if they have more than 2 active
pumps. The CTE (constant time estimate) flow calculation model assumes a
lead/lag pump configuration. Stations with 3 or 4 active pumps use more
complex control strategies (e.g., lead/lag/standby, duty/assist/jockey)
that require a different flow estimation approach. These stations require
manual configuration in the AF.

## Step 3: AF Hierarchy Structure

`export_af()` builds 6 rows per station in this hierarchy:

```
WASTEWATER
  \-- EST FLOW
        \-- {BASIN}
              \-- {STATION_NAME}          Element, Template: _Liftstation
                    |-- CMMS Asset ID     Attribute, Value: {CMMS_ASSET_ID}
                    |-- PI Tag            Attribute, Value: {PI_TAG}
                    |-- Wetwell Area      Attribute, Value: {Dia}
                    |-- cte               Element, Template: _cte_lead or _cte_leadsp
                    \-- hrt               Element, Template: _hrt
```

The `WASTEWATER\EST FLOW` root element and basin-level elements are
created once. Station elements are created per row in the import data.
