---
title: "Usage Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Usage Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

afbuildr provides a three-step pipeline for building, validating, and
formatting PI tags for wastewater lift stations:

| Step | Function | Input | Output |
|------|----------|-------|--------|
| 1 | `validate_gis()` | GIS data + PI Web API | `data/pi_tag_master_list.csv` |
| 2 | `build_af()` | master list + SCADA history | `data/pi_af_import.csv` |
| 3 | `export_af()` | AF import data | `xlsx/pi_af_loader.csv` |

Each step must complete before the next can run. The final output is a CSV
file formatted for the PI Asset Framework bulk import tool.

## Running the Pipeline

### Step 1: Build the PI Tag Master List

```{r}
library(afbuildr)
validate_gis()
```

This will:

1. Load lift station records from GIS (in-service stations with a CMMS asset
   ID)
2. Classify each station's PI tag as valid, missing, malformed, or
   unrecognized
3. Print a tree-formatted summary showing how many tags fall into each
   category
4. Attempt to guess correct tag names for missing or invalid tags by trying
   address-based candidates against the PI Web API
5. Export the results to `data/pi_tag_master_list.csv`
6. Print warnings for any duplicate CMMS asset IDs or PI tags

**Expected runtime:** Several minutes, depending on the number of tags that
need PI Web API validation. Progress messages are printed every 50 tags
(classification) and every 10 assets (guessing).

**What to check:** Review the tree summary output. A high percentage of
"Bad Tag" or "Missing Data (NA)" entries may indicate a GIS data quality
issue or PI Web API connectivity problem.

### Step 2: Check Pump Tags and Build AF Import Data

```{r}
build_af()
```

This will:

1. Read `data/pi_tag_master_list.csv` and filter to `WWL:` tags
2. Validate all pump tag combinations (4 pumps x 4 types = 16 per station)
3. Pull 7 days of SCADA history for the valid tags
4. Identify stations with VFD control or more than 2 pumps
5. Exclude stations with more than 2 pumps (these require manual handling)
6. Check for lead setpoint (`_WLLEADSP`) tags to determine the AF template
7. Join plant/basin assignments and pipe diameter reference data
8. Export to `data/pi_af_import.csv` and open in the RStudio viewer

**Expected runtime:** Longer than Step 1 due to the high number of pump tag
validations (16 per station) and the SCADA history retrieval. Progress
messages are printed every 100 tags.

**What to check:** The viewer will display the final import table. Verify
that the `BASIN`, `TEMPLATE`, and `Dia` columns look correct. Stations with
`_cte_lead` template do not have a lead setpoint tag; stations with
`_cte_leadsp` do.

### Step 3: Format for the PI AF CSV Loader

```{r}
export_af()
```

This will:

1. Read `data/pi_af_import.csv`
2. Build the hierarchical CSV structure required by the PI AF bulk import
   tool (6 rows per station)
3. Export to `xlsx/pi_af_loader.csv`

**Expected runtime:** Seconds (no API calls).

**What to check:** Open `xlsx/pi_af_loader.csv` and verify the AF hierarchy
looks correct before loading it into PI System Explorer.

## Running Individual Functions

You don't have to run the full pipeline. Individual functions can be called
for targeted tasks:

### Validate a specific set of tags

```{r}
# Build a data frame with the required columns
my_stations <- data.frame(
  STREET_NAME = c("MAIN", "OAK"),
  STREET_NUMBER = c("100", "200"),
  CMMS_ASSET_ID = c("LS001", "LS002"),
  PI_TAG = c("wwl:north:mai100_", "wwl:east:oak200_")
)

# Classify and validate
results <- classify_tags(my_stations)
print_tag_summary(results)
```

### Check pump tags for specific stations

```{r}
tags <- check_all_pump_tags(my_stations)
```

### Verify PI Web API connectivity

```{r}
# Should return a web ID string; NULL means the tag was not found
safe_get_webid("wwl:north:mai100_P1RUN")
```

## Output Files

| File | Created by | Description |
|------|-----------|-------------|
| `data/pi_tag_master_list.csv` | `validate_gis()` | All lift stations with validated/guessed PI tags |
| `data/pi_af_import.csv` | `build_af()` | Stations ready for AF import with basin, template, and diameter |
| `xlsx/pi_af_loader.csv` | `export_af()` | Final hierarchical CSV for PI AF bulk import tool |

## Troubleshooting

### "Error in get_webid(...)" or all tags show as bad

The PI Web API is unreachable or credentials are not configured. Verify your
`piwebapi` package configuration and network access. See
`vignette("technical-reference")` for setup details.

### Step 2 takes extremely long

Each station generates 16 pump tag candidates, each requiring a PI Web API
call. For 200 stations, that's 3,200 API calls. This is expected. If it
seems stuck, check the progress messages --- they update every 100 tags.

### "Missing data/ls_by_plant.RData" or "Missing data/dia_by_ls.RData"

These reference data files must be present before running Step 2. See
`vignette("technical-reference")` for details on how to obtain or
regenerate them.

### Duplicate warnings after Step 1

`validate_gis()` prints warnings for duplicate CMMS asset IDs and
duplicate PI tags. A small number of duplicates may be expected (e.g.,
stations with multiple GIS records). Large numbers suggest a data quality
issue in GIS.
